Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia

### 2024-11-27 analysis/analysis.qmd
Tidify model fitting and prediction
Leaving tuning out for now due to performance cost

```{r dependencies}
library("dplyr")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("parsnip")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("rgbif")
library("rpaleoclim")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidyr")

@@

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    #include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  # group_by(include) |>
  filter(n_occ > 50) |>
  arrange(-n_present) |>
  gt() |>
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

@@

```{r generate-background}
# TODO: maybe defer this to just before the model is actually fit ?
#   (and discard the background sample)
n_background <- 10000

flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE),
    bg_data = map(1:n(), \(i) { 
      background_sample(
        region = w_asia_land, 
        N = n_background,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE
      )
    }),
    occ = map2(occ, bg_data, bind_rows)
  ) |>
  select(-bg_data) ->
  flora
```

@@

```{r training-climate}
flora |> 
  mutate(
    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x))))
  ) ->
  flora
```

@@

```{r model}
rf_predictors <- function() {
  c(any_of("present"), starts_with("bio"))
}

# Drop any rows with NA values in any of the predictor variables
# TODO: Where do these actually come from? There's usually only a handful.
# occ <- drop_na(occ, rf_predictors())

# Model
# Parameters adapted from the example of a down-sampled RF model given in
#   https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
# TODO: tune?

rf_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger",
    probability = TRUE,
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
  )

# Fit to each taxon
# TODO: parallelise
flora |>
  mutate(
    fit = map(occ, \(occ) {
      fit(rf_model, factor(present) ~ ., data = select(occ, rf_predictors()))
    }, .progress = TRUE)
  ) ->
  flora
```

The output of the model is probalistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
This also makes it possible to analyses the predictions together as an assemblage.

```{r thresholding}
# TODO: thresholding
# - Pick a threshold
# - Reclassify pred raster
# - Extract predicted classification (for cur) for occ
# - ...or predict back onto the occurrence data? do we actually need the rasters for this?
# - Calculate sensitivity and with yardstick
# - Sum (SSS)
# - (Plot?)
# - Pick max SSS
```

@@

@@

```{r predict}
# TODO: tidy
rf_fit <- filter(flora, taxon == "Triticum turgidum dicoccum") |>
  pull(fit)
rf_fit <- rf_fit[[1]]

# TODO: Maybe combine stars and collapse period into dimension first?
rf_pred <- map(climate, predict, model = rf_fit, type = "prob")
rf_pred <- map(rf_pred, \(x) x[".pred_TRUE"])
rf_pred <- do.call(c, rf_pred)
rf_pred <- st_redimension(rf_pred, name = "period")

# Order and relabel periods
climate_periods <- rf_pred |>
  st_get_dimension_values("period") |>
  ordered(levels = c("LGM", "HS1", "BA", "YDS", "EH", "cur")) |>
  recode(
    "LGM" = "LGM (21 ka)",
    "HS1" = "HS1 (17.0-14.7 ka)",
    "BA"  = "BA (14.7-12.9 ka)",
    "YDS" = "YDS (12.9-11.7 ka)",
    "EH"  = "EH (11.7-8.326 ka)",
    "cur" = "current"
  )

rf_pred <- st_set_dimensions(rf_pred, "period", values = climate_periods)

ggplot() + 
  geom_stars(data = rf_pred) +
  facet_wrap(~period, ncol = 3) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = latlong) +
  labs(
    title = "Triticum turgidum dicoccum",
    fill = NULL,
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "italic")
  )
```

### Model assessment & verification

```{r todo}
# TODO: model performance assessment:
# - R2
# - Boyce index
# - AUC, ROC, AUC-ROC, AUC-PRG ?
# - TSS

>>>>

```{r dependencies}
library("cowplot")
library("dplyr")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("rpaleoclim")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

```{r tbl-occ-count}
#| tbl-cap: "Species modelled in this study."
# TODO: add to caption:
# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
# TODO: add count of species included?
flora |>
  mutate(
    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
    #include = n_occ > 50
  ) |>
  select(-taxon_original) |>
  # group_by(include) |>
  filter(n_occ > 50) |>
  arrange(-n_present) |>
  gt() |>
  # TODO: move synonyms to footnotes?
  cols_merge(
    c(taxon, taxon_detail),
    rows = !is.na(taxon_detail),
    pattern = "{1}<br><small>(incl. {2})</small>"
  ) |>
  fmt_percent(p_present, decimals = 0) |>
  cols_merge_n_pct(n_present, p_present) |>
  tab_style(cell_text(style = "italic"), cells_body(taxon))
```

@@

```{r generate-background}
# TODO: maybe defer this to just before the model is actually fit ?
#   (and discard the background sample)
# using a custom recipe step even?
n_background <- 10000

flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg_data = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_asia_land, n = n_background),
    occ = map2(occ, bg_data, bind_rows)
  ) |>
  select(-bg_data)
```

@@

```{r training-climate}
flora |> 
  mutate(
    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
    # TODO: where is the extra geometry column coming from?
    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
  ) ->
  flora
```

@@

```{r prepare-niche-model-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

# Data preprocessing recipe
# Following recommended preprocessing steps for RF classification from:
#   https://www.tmwr.org/pre-proc-table.html
niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
  update_role(gbif_key:latitude, new_role = "ID") |>
  update_role(present, new_role = "outcome") |>
  update_role(starts_with("bio"), new_role = "predictor") |>
  step_nzv(all_predictors())
```

```{r niche-models}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    probability = TRUE
  )

# TODO: Or... tune parameters? But this is very compute-intensive.
# niche_model <- rand_forest(
#   mode = "classification", 
#   mtry = tune(),
#   trees = tune(),
#   min_n = tune()
# ) |>
#   set_engine("ranger", probability = TRUE)

niche_workflow <- workflow(niche_recipe, niche_model)
```

We follow @ValaviEtAl2021a in using down-sampling and their recommended hyperparameters to avoid overfitting.

<!-- TODO: OR:
Hyperparameter tuning is important to avoid over-fitting [@CITE].
We tuned each model separately to maximise ROC AUC.
-->

```{r tune-niche-models}
# Tune model (separately for each taxon)
# TODO: would be nice but is extremely slow
# flora |>
#   slice_sample(n = 10) |>
#   mutate(
#     niche_tuning_grid = map(occ_split, function(data, recipe) {
#       grid_regular(
#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
#         trees(),
#         min_n(),
#         levels = 5
#       )
#     }, recipe = niche_recipe, .progress = TRUE),
#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
#       tune_grid(wflow, folds, grid = grid)
#     }, wflow = niche_workflow, .progress = TRUE),
#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
#     niche_model = map(niche_hparams, function(params, wflow) {
#       finalize_workflow(wflow, params)
#     }, wflow = niche_workflow)
#   )
```

```{r fit-niche-models}
# Fit best model to training data
flora <- flora |>
  mutate(
    niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
  )
```

The output of the model is probabilistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
This also makes it possible to analyses the predictions together as an assemblage.

```{r thresholding}
# TODO: thresholding?
# - Pick a threshold
# - Reclassify pred raster
# - Extract predicted classification (for cur) for occ
# - ...or predict back onto the occurrence data? do we actually need the rasters for this?
# - Calculate sensitivity and with yardstick
# - Sum (SSS)
# - (Plot?)
# - Pick max SSS
#
# Or just stick with probabilities?
```

@@

### Model assessment

```{r niche_model_test}
# Predict model on test data and assess performance
# TODO: model performance assessments:
# - [ ] R2
# - [ ] Boyce index
# - [x] AUC, ROC, AUC-ROC, AUC-PRG ?
# - [ ] TSS
# flora <- flora |>
#   mutate(
#     niche_test = map2(niche, occ_split, \(x, y) augment(x, testing(y))),
#     niche_test_roc_auc = map_dbl(niche_test, function(x) {
#       roc_auc(x, present, .pred_TRUE)$.estimate
#     }),
#     niche_test_accuracy = map_dbl(niche_test, function(x) {
#       accuracy(x, present, .pred_class)$.estimate
#     })
#   )

# TODO: filter to minimum accuracy? 0.85? 0.9? 0.95?
```

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  mutate(niche_test_fit = map(niche, collect_metrics)) |>
  unnest(niche_test_fit) |>
  filter(.metric == "accuracy") |> # TODO: or roc_auc?
  mutate(label = if_else(.estimate < 0.8, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_half_open() 
```

@@

```{r predict}
# TODO: structure this way from the beginning?
# Thanks to https://stackoverflow.com/questions/65183306/stars-package-how-to-define-additional-dimensions-based-on-an-attribute-filena
climate_pred <- do.call(c, map(climate, st_redimension, name = "variable")) |>
  st_redimension(name = "period") |>
  split("variable")

# TODO: very high memory usage
flora <- flora |>
  mutate(
    prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)), 
                     obj = climate_pred)
  )
```

```{r fig-niche-prediction}
niche_prediction <- filter(flora, taxon == "Linum bienne")$prediction[[1]]

# Order and relabel periods
climate_periods <- niche_prediction |>
  st_get_dimension_values("period") |>
  ordered(levels = c("LGM", "HS1", "BA", "YDS", "EH", "cur")) |>
  recode(
    "LGM" = "LGM (21 ka)",
    "HS1" = "HS1 (17.0-14.7 ka)",
    "BA"  = "BA (14.7-12.9 ka)",
    "YDS" = "YDS (12.9-11.7 ka)",
    "EH"  = "EH (11.7-8.3 ka)",
    "cur" = "current"
  )

niche_prediction <- st_set_dimensions(niche_prediction, "period", values = climate_periods)

ggplot() + 
  geom_stars(data = select(niche_prediction, prediction.TRUE.) > 0.5) +
  facet_wrap(~period, ncol = 3) +
  # scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = latlong) +
  labs(
    title = "Linum bienne",
    fill = NULL,
    x = NULL,
    y = NULL
  ) +
  theme_map() +
  theme(
    plot.title = element_text(face = "italic")
  )
```

### Comparison to archaeological occurrences

```{r sum-pred-eh}
flora |>
  mutate(pred_eh = map(fit, \(x) predict(climate$EH, model = x, type = "prob"))) ->
  x

pred_eh <- x$pred_eh
names(pred_eh) <- x$taxon
pred_eh <- do.call(c, map(pred_eh, select, .pred_TRUE))
pred_eh <- merge(pred_eh, name = "taxon")
sum_pred_eh <- st_apply(pred_eh, c("x", "y"), sum, na.rm = TRUE)
```

```{r fig-sum-pred-eh}
#| fig-cap: predicted species richness for the early Holocene
ggplot() + 
  geom_stars(data = sum_pred_eh) +
  scale_fill_batlow(reverse = FALSE) +
  coord_sf(crs = latlong) +
  labs(
    title = "Predicted species richness",
    subtitle = "Early Holocene (11.7–8.326 ka)",
    fill = NULL,
    x = NULL,
    y = NULL
  )

saveRDS(pred_eh, "pred_eh.Rdata")
```

```{r fig-pred-observed-eh}
